---
title: "Using OPCC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using OPCC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

OPCC links Ontario postal codes to Statistics Canada 2021 census geographies
through open, redistributable evidence.

**Two access paths:**

| You want to... | Go to |
|---|---|
| Look up postal codes or join census geographies to your own data | [Part 1](#part-1-use-opcc-with-your-data) -- uses pre-built artifacts downloaded automatically by the package; no source checkout needed |
| Verify artifacts or reproduce the DA roll-up from the DB artifact | [Part 2](#part-2-reproduce-and-verify-artifacts) -- uses package functions only; no source checkout needed |
| Rebuild the DB correspondence from raw Statistics Canada sources | [Rebuild M2](#rebuild-m2-from-raw-sources-maintainers-only) -- requires a source checkout, ~1.6 GB of public inputs, and `sf`; maintainer workflow |

## Part 1: Use OPCC with your data

### Install

Requires R >= 4.1.

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("lennon-li/OPCC")
```

### Key concepts

OPCC links postal codes to two Statistics Canada census geographies:

- **DB (Dissemination Block):** the smallest census unit, typically a city
  block or rural area. Identified by `DBUID`.
- **DA (Dissemination Area):** a group of DBs, roughly 400-700 persons.
  Identified by `DAUID`. Every DB belongs to exactly one DA.

Each postal code can map to multiple geographies. OPCC retains all defensible
links with an `allocation_weight` (the share of observed address evidence for
that link). The `best_link` flag marks the single strongest link per postal
code. Unmatched postal codes are reported, never fabricated.

### Normalize postal codes

All OPCC functions normalize internally, but you can normalize explicitly:

```{r, eval = FALSE}
library(OPCC)

normalize_postal_code(c("m5v3a8", "K1A-0A6", "k1a0a6"))
#> [1] "M5V 3A8" "K1A 0A6" "K1A 0A6"
```

### Look up a single postal code

The first call downloads and caches the artifact (~15 MB). Subsequent calls
use the local cache.

```{r, eval = FALSE}
# Best dissemination area (DA) link
pc_to_geo("M5V 3A8", level = "DA", all_links = FALSE)

# Every defensible DA link with allocation weights
pc_to_geo("M5V 3A8", level = "DA")

# Best dissemination block (DB) link
pc_to_geo("M5V 3A8", level = "DB", all_links = FALSE)

# Every DB link
pc_to_geo("M5V 3A8", level = "DB")
```

Unmatched postal codes are not silently dropped. They appear in the `unmatched`
attribute of the result:

```{r, eval = FALSE}
result <- pc_to_geo(c("M5V 3A8", "ZZZ 9Z9"), level = "DA")
attr(result, "unmatched")
#> [1] "ZZZ 9Z9"
```

### Join OPCC geographies to your own data

The most common workflow: you have a data frame with a postal-code column and
you want census geography identifiers.

```{r, eval = FALSE}
library(OPCC)

my_data <- data.frame(
  id = 1:5,
  postal_code = c("M5V 3A8", "K1A 0A6", "N6A 1B1", "P7A 1A1", "ZZZ 9Z9"),
  value = c(10, 20, 30, 40, 50),
  stringsAsFactors = FALSE
)

# --- DA join (one best DA per postal code) ---

da <- get_da_correspondence(vintage = "2026-07-20")
da_best <- da[da$best_link, c("postal_code", "DAUID", "allocation_weight")]

merged_da <- merge(
  my_data,
  da_best,
  by = "postal_code",
  all.x = TRUE
)
# Rows with unmatched postal codes keep NA for DAUID and allocation_weight.

# --- DB join (one best DB per postal code) ---

db <- get_correspondence(vintage = "2026-07-19-geonames-amendment")
db_best <- db[db$best_link, c("postal_code", "DBUID", "DAUID", "allocation_weight")]

merged_db <- merge(
  my_data,
  db_best,
  by = "postal_code",
  all.x = TRUE
)

# --- Keep all many-to-many links ---
# Skip the best_link filter to retain every candidate geography and its
# allocation weight. Your join will produce multiple rows per postal code.

merged_all <- merge(my_data, da, by = "postal_code", all.x = TRUE)
```

**Columns returned by the join:**

| Column | Meaning |
|---|---|
| `postal_code` | Normalized `A1A 1A1` postal code |
| `DBUID` | 2021 Dissemination Block identifier (DB artifact) |
| `DAUID` | 2021 Dissemination Area identifier (both artifacts) |
| `allocation_weight` | Share of observed address evidence for this link; sums to 1 per postal code |
| `best_link` | `TRUE` for the single strongest link per postal code |
| `n_contributing_dbs` | Number of DBs contributing to a DA link (DA artifact) |
| `contributing_dbuids` | Pipe-delimited DBUID list behind a DA link (DA artifact) |
| `source_vintages` | Source evidence vintage(s) behind the link |

With `dplyr`:

```{r, eval = FALSE}
library(dplyr)

da <- get_da_correspondence(vintage = "2026-07-20")

my_data |>
  mutate(postal_code = normalize_postal_code(postal_code)) |>
  left_join(
    da |> filter(best_link) |> select(postal_code, DAUID, allocation_weight),
    by = "postal_code"
  )
```

### Download and cache artifacts

`get_correspondence()` and `get_da_correspondence()` download the versioned
artifact, verify its SHA-256 checksum, cache it locally, and return a data
frame. Subsequent calls use the cache.

By default, artifacts are cached in a persistent user directory
(`tools::R_user_dir("OPCC", "cache")`), so they survive across R sessions.
You can override this with `cache_dir`.

```{r, eval = FALSE}
# List available release vintages
list_vintages(level = "DB")
list_vintages(level = "DA")

# Download using the default persistent cache
db <- get_correspondence(vintage = "2026-07-19-geonames-amendment")
da <- get_da_correspondence(vintage = "2026-07-20")

# Or use an explicit cache directory
cache <- file.path(tempdir(), "opcc-cache")
db <- get_correspondence(
  vintage = "2026-07-19-geonames-amendment",
  cache_dir = cache
)

# Pass a pre-loaded artifact to skip re-reading
pc_to_geo("M5V 3A8", level = "DA", correspondence = da)
pc_to_geo("M5V 3A8", level = "DB", correspondence = db)
```

### Verify a release

```{r, eval = FALSE}
# Download, checksum, and validate schema and invariants
validate_release(vintage = "2026-07-19-geonames-amendment", level = "DB",
                 cache_dir = cache)
validate_release(vintage = "2026-07-20", level = "DA", cache_dir = cache)

# Inspect provenance metadata
release_manifest(vintage = "2026-07-20", level = "DA", cache_dir = cache)
```

### Offline and air-gapped use

After caching a verified release once, use `offline = TRUE` to require the
local cache and block network access:

```{r, eval = FALSE}
validate_release(vintage = "2026-07-20", level = "DA",
                 cache_dir = cache, offline = TRUE)
da <- get_da_correspondence(vintage = "2026-07-20",
                            cache_dir = cache, offline = TRUE)
```

You can also supply a local artifact file directly to `pc_to_point()`:

```{r, eval = FALSE}
pts <- pc_to_point("K0A 0A1",
                   point_file = "/path/to/opcc_m1_geonames_points.csv.gz")
```

### Source-qualified point lookups

`pc_to_point()` returns supplementary GeoNames point observations. These are
source-labeled and separate from NAR address evidence:

```{r, eval = FALSE}
pc_to_point("K0A 0A1")
pc_to_point("K0A 0A1", source = "geonames")
```

### Import your own open evidence locally

Use this only for redistributable, open data. Local layers stay separate from
canonical OPCC releases. The API rejects Canada Post, PCCF, and PCCF+ inputs.

```{r, eval = FALSE}
my_source <- utils::read.csv("/path/to/municipal-postal-data.csv",
                             stringsAsFactors = FALSE)

adapter <- new_source_adapter(
  source_id = "municipal_registry",
  licence = "Open Government Licence - Municipality",
  lineage = "Municipal open address registry",
  retrieval_date = "2026-07-20",
  schema_map = list(
    postal_code = "postal",
    latitude = "lat",
    longitude = "lon"
  ),
  endpoint = "https://example.org/open-data",
  checksum = digest::digest("/path/to/municipal-postal-data.csv",
                            algo = "sha256", file = TRUE)
)

layer <- build_source_layer(my_source, adapter, on_invalid = "quarantine")
profile_source_layer(layer)

bundle <- contribution_bundle(layer, output_dir = "local-contributions",
                              fixture_rows = 100L)
contribution_issue_url(bundle)
```

## Part 2: Reproduce and verify artifacts

Every published artifact can be downloaded, verified, and reproduced using
package functions alone. No source checkout is needed.

### Verify published artifacts

Download each artifact, check its SHA-256 against the release index, and
validate schema and correspondence invariants (unique keys, weight sums,
one best link per postal code):

```{r, eval = FALSE}
validate_release(vintage = "2026-07-19-geonames-amendment", level = "DB")
validate_release(vintage = "2026-07-20", level = "DA")
```

Inspect the full provenance metadata (source checksums, code version, build
timestamp, row counts):

```{r, eval = FALSE}
release_manifest(vintage = "2026-07-19-geonames-amendment", level = "DB")
release_manifest(vintage = "2026-07-20", level = "DA")
```

### Reproduce the DA artifact from the DB artifact

The DA correspondence (M5) is a deterministic attribute roll-up of the DB
correspondence (M2). `aggregate_da_correspondence()` performs the same
computation as the M5 build script: it sums DB allocation weights within
each DA, preserves contributing-DB lineage, and selects the best DA link.

```{r, eval = FALSE}
db <- get_correspondence(vintage = "2026-07-19-geonames-amendment")
da_reproduced <- aggregate_da_correspondence(db)

# Compare with the published DA artifact
da_published <- get_da_correspondence(vintage = "2026-07-20")
stopifnot(identical(nrow(da_reproduced), nrow(da_published)))
stopifnot(identical(
  sort(unique(da_reproduced$postal_code)),
  sort(unique(da_published$postal_code))
))
```

### Reproduce offline

After caching artifacts once, all verification and reproduction works without
network access:

```{r, eval = FALSE}
cache <- tools::R_user_dir("OPCC", "cache")

validate_release(vintage = "2026-07-19-geonames-amendment", level = "DB",
                 cache_dir = cache, offline = TRUE)
validate_release(vintage = "2026-07-20", level = "DA",
                 cache_dir = cache, offline = TRUE)

db <- get_correspondence(vintage = "2026-07-19-geonames-amendment",
                         cache_dir = cache, offline = TRUE)
da <- aggregate_da_correspondence(db)
```

### Rebuild M2 from raw sources (maintainers only)

Rebuilding the DB correspondence from the original Statistics Canada NAR
release requires a source checkout, ~1.6 GB of public inputs, and the `sf`
package (which needs system libraries GDAL, GEOS, and PROJ). This is a
maintainer or auditor workflow; the published artifact is immutable and
checksum-verified.

On Ubuntu: `sudo apt-get install libgdal-dev libgeos-dev libproj-dev`.
On macOS: `brew install gdal geos proj`. On Windows: install
[Rtools](https://cran.r-project.org/bin/windows/Rtools/) and use the
`sf` binary from CRAN.

```{r, eval = FALSE}
packages <- c("digest", "dplyr", "jsonlite", "readr", "sf")
missing <- packages[!vapply(packages, requireNamespace, logical(1),
                            quietly = TRUE)]
if (length(missing)) install.packages(missing)
```

```bash
git clone https://github.com/lennon-li/OPCC.git
cd OPCC
```

All large inputs go to `.scratch/` (git-ignored) and must not be committed.

**Stage 1: M1 -- centroids and DB assignment.**
Download and profile the NAR and GeoNames reference layer:

```bash
Rscript scripts/m1_nar_profile.R
Rscript scripts/m4_geonames_profile.R
Rscript scripts/m1_build_centroids.R
```

Download these three Statistics Canada files and extract them to the paths
shown:

| File | URL | Extract to |
|---|---|---|
| Province/territory boundary | `https://www12.statcan.gc.ca/census-recensement/2021/geo/sip-pis/boundary-limites/files-fichiers/lpr_000b21a_e.zip` | `.scratch/shp/` |
| DB boundary | `https://www12.statcan.gc.ca/census-recensement/2021/geo/sip-pis/boundary-limites/files-fichiers/ldb_000b21a_e.zip` | `.scratch/shp/` |
| Geographic Attribute File | `https://www12.statcan.gc.ca/census-recensement/2021/geo/aip-pia/attribute-attribs/files-fichiers/2021_92-151_X.zip` | `.scratch/gaf/` |

After extraction, these paths must exist:

```text
.scratch/shp/lpr_000b21a_e.shp
.scratch/shp/ldb_000b21a_e.shp
.scratch/gaf/2021_92-151_X.csv
```

Assign centroids to dissemination blocks and attach GAF geography columns:

```bash
Rscript scripts/m1_gaf_rollup.R
```

**Stage 2: M2 -- postal-code-to-DB correspondence.**

```bash
Rscript scripts/m2_build_correspondence.R
```

Output is written to `.scratch/m2/`:

- `m2_correspondence.csv` -- one row per postal-code/DBUID pair
- `m2_manifest.json` -- source checksums, code version, build time, row counts

Verify the build invariants:

```{r, eval = FALSE}
candidate <- utils::read.csv(".scratch/m2/m2_correspondence.csv",
                             stringsAsFactors = FALSE)
stopifnot(!anyDuplicated(candidate[c("postal_code", "DBUID")]))
weights <- tapply(candidate$allocation_weight, candidate$postal_code, sum)
stopifnot(all(abs(weights - 1) < 1e-8))
stopifnot(all(tapply(candidate$best_link, candidate$postal_code, sum) == 1L))
```

**Exact byte-level rebuild.**
To reproduce the published artifact bytes exactly, check out the producer
revision recorded in the published manifest rather than the tip of `main`.
See `docs/m2-reproduction.md` and `docs/m5-reproduction.md` for the pinned
revisions, expected SHA-256 values, and full verification sequences.
