---
title: "Using OPCC reproducibly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using OPCC reproducibly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

OPCC keeps the release vintage and evidence source visible. Normalize input
first; the default geography lookup returns every allocation link.

```{r, eval = FALSE}
library(OPCC)
pc_to_geo("M5V3A8")
pc_to_geo("M5V3A8", all_links = FALSE)
```

## Download and use the DB and DA artifacts directly

`get_correspondence()` downloads the versioned postal-code-to-DB CSV artifact,
verifies its SHA-256 checksum, caches it, and returns the artifact as a data
frame. `get_da_correspondence()` does the same for the direct postal-code-to-DA
artifact. Keep all links unless a workflow explicitly requires the deterministic
best link.

```{r, eval = FALSE}
cache <- file.path(tempdir(), "opcc-cache")

# Postal code -> dissemination block (DB) artifact.
db <- get_correspondence(
  vintage = "2026-07-19-geonames-amendment",
  cache_dir = cache
)
db_links <- pc_to_geo("M5V 3A8", level = "DB", correspondence = db)
db_best <- pc_to_geo("M5V 3A8", level = "DB", all_links = FALSE,
                     correspondence = db)

# Postal code -> dissemination area (DA) artifact, with contributing DB lineage.
da <- get_da_correspondence(vintage = "2026-07-20", cache_dir = cache)
da_links <- pc_to_geo("M5V 3A8", level = "DA", correspondence = da)
da_best <- pc_to_geo("M5V 3A8", level = "DA", all_links = FALSE,
                     correspondence = da)
```

The returned `db` and `da` objects are the verified artifact tables. DB rows
carry `DBUID`, `DAUID`, source fields, allocation weight, and `best_link`; DA
rows carry `DAUID`, allocation weight, `best_link`, and contributing-DB
lineage. Unmatched normalized postal codes are retained in
`attr(db_links, "unmatched")` or `attr(da_links, "unmatched")`.

## Reproduce a verified artifact retrieval

The following exact sequence retrieves each immutable, commit-pinned artifact,
verifies its artifact and manifest checksums, and then repeats without network
access from the verified cache.

```{r, eval = FALSE}
cache <- file.path(tempdir(), "opcc-repro-cache")

validate_release(
  vintage = "2026-07-19-geonames-amendment",
  level = "DB",
  cache_dir = cache
)
validate_release(vintage = "2026-07-20", level = "DA", cache_dir = cache)

# Inspect the provenance that was checksum-verified above.
db_manifest <- release_manifest(
  vintage = "2026-07-19-geonames-amendment", level = "DB", cache_dir = cache
)
da_manifest <- release_manifest(vintage = "2026-07-20", level = "DA", cache_dir = cache)

# The same calls now use only the verified local cache.
validate_release(
  vintage = "2026-07-19-geonames-amendment",
  level = "DB", cache_dir = cache, offline = TRUE
)
validate_release(
  vintage = "2026-07-20", level = "DA", cache_dir = cache, offline = TRUE
)
```

This reproduces verified use of the published artifacts. Rebuilding the
producer artifacts from raw source data is a separate maintainer workflow:
follow `docs/m2-reproduction.md` for DB evidence and
`docs/m5-reproduction.md` for the deterministic DB-to-DA roll-up. Those guides
pin source inputs, producer revisions, output paths, and expected hashes.

Use `offline = TRUE` in reproducible or air-gapped environments after caching a
verified release. `validate_release()` checks the cached artifact's checksum,
postal-code/DB uniqueness, allocation-weight sums, and best-link invariant.

```{r, eval = FALSE}
validate_release(offline = TRUE)
release_manifest(offline = TRUE)
```

`pc_to_point()` provides GeoNames-derived supplementary point records. These
are explicitly source-labeled and must not be interpreted as NAR address
evidence.

```{r, eval = FALSE}
pc_to_point("K0A0A1")
```
