---
title: "Using OPCC reproducibly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using OPCC reproducibly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

OPCC keeps the release vintage and evidence source visible. Normalize input
first; the default geography lookup returns every allocation link.

```{r, eval = FALSE}
library(OPCC)
pc_to_geo("K1A0A6")
pc_to_geo("K1A0A6", all_links = FALSE)
```

Use `offline = TRUE` in reproducible or air-gapped environments after caching a
verified release. `validate_release()` checks the cached artifact's checksum,
postal-code/DB uniqueness, allocation-weight sums, and best-link invariant.

```{r, eval = FALSE}
validate_release(offline = TRUE)
release_manifest(offline = TRUE)
```

`pc_to_point()` provides GeoNames-derived supplementary point records. These
are explicitly source-labeled and must not be interpreted as NAR address
evidence.

```{r, eval = FALSE}
pc_to_point("K0A0A1")
```
