---
title: "Reproduce and verify the M2 correspondence artifact"
format:
  html:
    toc: true
execute:
  eval: false
---

## Purpose and scope

This guide has two reproducibility paths:

1. Verify the published 2026-06-26 M2 baseline exactly, using its public,
   commit-pinned files and SHA-256 checksums.
2. Rebuild the current correspondence candidate from its public source inputs.

The published baseline is a frozen NAR-only artifact with 414,207
postal-code/DBUID rows and 282,409 postal codes. The current build script also
adds separately labeled GeoNames point-in-polygon evidence. Its output is a
newer candidate, not a byte-for-byte replacement for the frozen baseline.

OPCC represents observed postal associations. It does not reconstruct Canada
Post assignments and must never use Canada Post, PCCF, or PCCF+ data.

## Verify the published artifact

Clone the repository and run this from its root. The remote check downloads
the public manifest and compressed CSV, verifies their SHA-256 values, then
checks unique postal-code/DBUID keys, allocation weights, and one `best_link`
per postal code.

```{r}
status <- system2(
  "Rscript",
  c("scripts/m3_validate_release.R", "--remote")
)
stopifnot(identical(status, 0L))
```

The expected result identifies the remote release endpoint and reports
414,207 rows and 282,409 postal codes. To verify the versioned files already
present in a source checkout instead, omit `"--remote"`.

```{r}
status <- system2("Rscript", "scripts/m3_validate_release.R")
stopifnot(identical(status, 0L))
```

## Rebuild the current candidate

The full build is intentionally separate from package use. It downloads and
creates large scratch inputs under `.scratch/`; these files are git-ignored and
must not be committed.

### Install R dependencies

```{r}
packages <- c("digest", "dplyr", "jsonlite", "readr", "sf")
missing <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) install.packages(missing)
```

### Obtain public inputs and generate the M1 intermediates

Download the Statistics Canada NAR release and the GeoNames reference layer.
The scripts record their provenance and reuse cached downloads when present.

```{r}
stopifnot(system2("Rscript", "scripts/m1_nar_profile.R") == 0L)
stopifnot(system2("Rscript", "scripts/m4_geonames_profile.R") == 0L)
stopifnot(system2("Rscript", "scripts/m1_build_centroids.R") == 0L)
```

Download the 2021 Statistics Canada DB boundary shapefile and Geographic
Attribute File (GAF) from the URLs in [M1 reproduction](m1-reproduction.md).
Extract them so these paths exist:

```text
.scratch/shp/ldb_000b21a_e.shp
.scratch/gaf/2021_92-151_X.csv
```

Then assign centroids to DBs and attach GAF geography columns.

```{r}
stopifnot(system2("Rscript", "scripts/m1_gaf_rollup.R") == 0L)
```

### Build and inspect the candidate

```{r}
stopifnot(system2("Rscript", "scripts/m2_build_correspondence.R") == 0L)

candidate <- utils::read.csv(
  ".scratch/m2/m2_correspondence.csv",
  stringsAsFactors = FALSE
)
manifest <- jsonlite::read_json(".scratch/m2/m2_manifest.json", simplifyVector = TRUE)

stopifnot(!anyDuplicated(candidate[c("postal_code", "DBUID")]))
weights <- tapply(candidate$allocation_weight, candidate$postal_code, sum)
stopifnot(all(abs(weights - 1) < 1e-8))
stopifnot(all(tapply(candidate$best_link, candidate$postal_code, sum) == 1L))

list(rows = nrow(candidate), postal_codes = length(unique(candidate$postal_code)), manifest = manifest)
```

The candidate and manifest are written to `.scratch/m2/`. Review their source
labels and manifest before proposing a new release. Do not overwrite the
versioned directory under `releases/m2/` or commit raw source data.

## Optional: import your own open evidence locally

Use this section only for redistributable, open evidence that you are allowed
to inspect and share. Local data stay separate from canonical OPCC releases.
The API rejects Canada Post, PCCF, and PCCF+ identifiers and provenance.

Create a CSV outside the repository or in an ignored directory. It needs a
postal-code column; coordinates are optional, but if supplied they must be
named `latitude` and `longitude` and be decimal degrees.

```{r}
# Run from a source checkout. This loads the local-layer API without installing
# the package first.
source("R/opcc.R")

my_data <- utils::read.csv("/path/to/municipal-postal-data.csv", stringsAsFactors = FALSE)

# Adapt column names once at import. Do not alter the original source file.
names(my_data)[names(my_data) == "postal"] <- "postal_code"
names(my_data)[names(my_data) == "lat"] <- "latitude"
names(my_data)[names(my_data) == "lon"] <- "longitude"

source_file <- "/path/to/municipal-postal-data.csv"
adapter <- new_source_adapter(
  source_id = "municipal_registry",
  licence = "Open Government Licence - Municipality",
  lineage = "Municipal open address registry",
  retrieval_date = "2026-07-20",
  schema_map = list(postal_code = "postal_code"),
  endpoint = "https://example.org/open-data",
  checksum = digest::digest(source_file, algo = "sha256", file = TRUE)
)

layer <- build_source_layer(my_data, adapter)
profile_source_layer(layer)
```

The layer has explicit local-only provenance and never changes a canonical
release. If the licence permits redistribution, create a small review bundle
rather than sharing the full source file.

```{r}
bundle <- contribution_bundle(
  layer,
  output_dir = "local-contributions",
  fixture_rows = 100L
)
bundle
```

Before contributing, review the licence, endpoint, lineage, checksum, fixture,
and quality report. See [M4 source adapters](m4-contributing-source.md) for
the maintainer review process.
